---
title: 'Week 4 Assignment: Health and Economic Impacts of Severe Weather'
author: "krishna"
date: "4/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

## Instructions

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.  
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

## Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. The data for this assignment can be downloaded from [here](Storm Data)  

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

Information on the Data: [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

## Synopsis

The goal of the assignment is to explore the NOAA Storm Database and explore the effects of severe weather events on both population and economy.The database covers the time period between 1950 and November 2011.

The following analysis answers the follwing two questions: 

  1. Across the United States, which types of events (as indicated in the ENVTYPE variable) are most harmful with respect to population health?  
  2. Across the United States, which types of events have the greatest economic consequences?  

## Loading and Preprocessing the data 

1. Environment setup

```{r read libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```

2.  Read in the data
```{r}
if (!file.exists('data2.csv.bz2')){
  download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2',
                destfile = paste0(getwd(), '/data2.csv.bz2'),
                method = 'curl', quiet = T)
}
raw <- read.csv('data2.csv.bz2', stringsAsFactors = F)
```

3. Inspect the data  
```{r inspect data}
names(raw)
```

4. Select data  

Since we're only interested in the correlation of the nation-wide event type to health and economic consequense, only the following columns will be selected for further analysis:  

- EVTYPE: event type  
- FATALITIES: number of fatalities  
- INJURIES: number of injuries  
- PROPDMG: property damage (dollars)  
- PROPDMGEXP: magnitude of property damage (K = thousands, M = millions, B = billions)  
- CROPDMG: crop damage (dollars)  
- CROPDMGEXP: magnitude of crop damage (H = hundreds, K = thousands, M = millions, B = billions)  

```{r}
selectCol <- c('EVTYPE', 'FATALITIES', 'INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')
rawSel <- raw[, selectCol]
summary(rawSel)
```






